---
title: "Reproducible Research: 1st Assignment"
output: html_document
---

This assignment is divided in steps, and the data analyzed was collected by an "activity monitoring device"  
during the months of October and November, 2012.  

## Loading and preprocessing the data  

Loading data and converting date to date type.

```{r}
activity <- read.csv("activity.csv", colClasses = c("numeric", "character", "numeric"))
activity$date <- as.Date(activity$date)
head(activity)
```

## What is mean total number of steps taken per day?  

- Because the data has NA values, I will omit them.

```{r}
act_1 <- na.omit(activity)
head(act_1)
```

- Calculating the total number of steps taken per day.

```{r}
Total_steps <- aggregate(steps ~ date, act_1, sum)
colnames(Total_steps) <- c("Date","Steps")
head(Total_steps)
```

- Making a histogram of the total number of steps taken each day.

```{r}
hist(Total_steps$Steps, main= "Total Steps per day", xlab="Steps",
     breaks= 20, col= "cyan")
```

- Calculate and report the mean and median of the total number of steps taken per day.
```{r}
Steps_mean <- mean(Total_steps$Steps); Steps_mean
Steps_median <- median(Total_steps$Steps); Steps_median
```

## What is the average daily activity pattern?

- Calculating the average.

```{r}
int_av <- aggregate(steps ~ interval, data = act_1, FUN = mean)
```

- Plotting

```{r}
plot(int_av, type = "l", xlab = "5 min interval",
     ylab = "Average across all days", main = "Average number of steps")
```

- Which interval, on average across all the days, has the max number of steps?

```{r}
max_int <- int_av[which.max(int_av$steps),]; max_int
```

So the 835th interval is the one with most steps with 206 steps.


## Imputing missing values

- Calculating the total number of missing values in the data set.

```{r}
sum(is.na(activity))
```

- Replacing the missing values by the mean of the 5 min interval .

```{r}
act_2 <- transform(activity, steps = ifelse(is.na(activity$steps), 
      int_av$steps[match(activity$interval, int_av$interval)], activity$steps))

head(act_2)
sum(is.na(act_2))
```

- So no Na values remain on the new data set

- Calculating the new total number of steps and making the new histogram.

```{r}
Total_steps_1 <- aggregate(steps ~ date, act_2, sum)
colnames(Total_steps_1) <- c("Date","Steps")
head(Total_steps_1)

hist(Total_steps_1$Steps, main= "Total Steps per day without Na values", 
     xlab="Steps", breaks= 20, col= "pink")
```

- Calculating the new mean and median for the data set with filled missing values.

```{r}
Steps_mean_1 <-mean(Total_steps_1$Steps); Steps_mean_1
Steps_median_1 <- median(Total_steps_1$Steps); Steps_median_1
```

- Comparing the new and old mean/median.

```{r}
mean_diff <- Steps_mean_1 - Steps_mean ; mean_diff
median_diff <- Steps_median_1 - Steps_median ; median_diff
```

- The mean has not changed, but the new median is slightly bigger than the old one. 

## Are there differences in activity patterns between weekdays and weekends?

- I'm going to use the act_2 data set with filled missing values.

- First I'll be using the weekdays function to create a new variable in the data frame.

```{r}
act_2$weekdays <- factor(format(act_2$date, "%A"))
levels(act_2$weekdays)
```

- And separating the days in weekday and weekend.

```{r}
levels(act_2$weekdays) <- list(weekday = c("Monday","Tuesday","Wednesday",
                              "Thursday","Friday"), weekend = c("Saturday","Sunday"))
levels(act_2$weekdays)
table(act_2$weekdays)
```

- Calculating the average of steps in both weekday and weekend periods.

```{r}
av_steps <- aggregate(act_2$steps,
                      list(interval = act_2$interval, weekdays = act_2$weekdays), 
                      FUN = "mean")
names(av_steps)[3] <- "mean"
head(av_steps)
```

- Changing

- Plotting the average of steps

```{r}
library(lattice)
xyplot(av_steps$mean ~ av_steps$interval | av_steps$weekdays,
       layout= c(1,2), type = "l",
       xlab= "Interval", ylab = "Number of Steps")
```

- Is evident on the plot that the activity on the weekend is different that on weekday,  
generally speaking the activity on the weekend is bigger than on the weekday.
